How do changes in ex vary between countries?

### 2022-03-07 draft_paper.Rmd
added additional desciptive comments on modelling strategy so far

---
title: "How do changes in ex vary between countries?"
author: "Jon Minton"
date: "22/02/2022"
output:
  html_document: default
  word_document: default
---

@@

@@

>>>>

---
title: "How do changes in ex vary between countries?"
author: "Jon Minton, Lucinda Hiam"
date: "22/02/2022"
output:
  html_document: default
  word_document: default
---

@@

```{r, cache=TRUE}

@@

### Joinpoint/Segmented regression approach

Though there may be systemic differences in average changes in life expectancy over time, such differences may not be patterned by decade, and so the above model specification may not have been appropriate to detect any such changes. Instead of looking for systemic differences by decade, we can instead look at whether there are any systemic and detectable 'breaks' in the trends over time. If our data were life expectancy over time, this could be achieved through joinpoint analysis, in which the slope is allowed to change over time. As instead are data are annual changes in life expectancy, we are interested in detecting changes in intercept over time. If we define our Null (no change) model as follows:

$$
M_0 : y(t) = \alpha + e
$$
Where $e \backsim N(0, \sigma^2)$. i.e. all observations $y$ are draws from $N(\alpha, \sigma^2)$. The alternative model specification against which this is compared is then:

$$
M_A : y(t) = \alpha_0 + \beta_1 T(t) + e 
$$
Where $T(t)$ is defined as $0$ where $t \lt \tau$ and $1$ where $t \geq \tau$. ($\alpha_0$ is used in $M_A$ to indicate that this parameter is not directly comparable with $\alpha$ in $M_0$).

This specification shows that there is an additional parameter $\tau$ to be determined in order to produce $M_A$. This $\tau$ parameter is the breakpoint in the series. The selection of $\tau$ will be determined by numeric optimisation, by selecting the value of $\tau$ that minimised AIC.

Once the best possible $M_A$, using the best value of $\tau$, is selected, $M_A$ will be compared with $M_0$ using an F-test, allowing us to determine whether there is convincing evidence of any specific break in the data.

We will start with a single country, starting age, and sex, and with manual selection of $\tau$, then generalise and automate further.

# Let's start with a single example population. (There are many different population subgroups, hence the 
# benefit of being able to automate this later.) We're going to start with England & Wales, females, life 
# expectancy at birth (x == 0)

example_df <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
example_df <- 
  example_df %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) # %>% 
  # ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
  # geom_line() +
  # geom_hline(yintercept = 0) + 
  # stat_smooth()

# The null model is really simple: just an intercept 
model_null <- lm(delta_ex ~ 1, data = example_df)
summary(model_null)
#

# The code below shows how we can be even simpler, and force the regression model not to 
# even have an intercept (i.e. have zero predictor variables)

# model_nuller <- lm(delta_ex ~ 0, data = example_df)
# summary(model_nuller)

# And we can compare the intercept with the no intercept model as follows 
# anova(model_nuller, model_null)
# a statistically significant p value shows the intercept is definitely worth including in the 
# model. Substantively that it's wrong to assume no improvement on average in ex over time since the 1980s

# There are lots of possible alternative models, some examples of which are below:
model_alt01 <- lm(delta_ex ~ year >= 1990, data = example_df)
summary(model_alt01)

model_alt02 <- lm(delta_ex ~ year >= 1995, data = example_df)
summary(model_alt02)

model_alt03 <- lm(delta_ex ~ year >= 1997, data = example_df)
summary(model_alt03)

model_alt04 <- lm(delta_ex ~ year >= 2008, data = example_df)
summary(model_alt04)

model_alt05 <- lm(delta_ex ~ year >= 2011, data = example_df)
summary(model_alt05)

# each of these models allows an adjustment to the horizontal line after a given year has been reached. 
# In the examples above some possible given years are 1990, 1995, 1997, 2008 and 2011. 
# Each of these is a candidate value for tau. But we've not looked at all possible/plausible values of this.

# Hence, it would be good to automate further to search for each possible tau and see how the fit and model 
# results change.

# We'll start to automate by building a function which creates T_param given tau, and runs and returns the model
# with this type of T_param
# (If tau is 1990, then T_param is equivalent to model_alt01 above; if tau were 1995, then it's equivalent to model_alt02, and so on. The only difference is that the new coefficient created has the same name, T_param, 
# whereas in the above it would have a different name each time, because the coefficient is being 
# created in situ, within the formula specification, rather than by pre-processing the dataframe)
run_alt_with_given_tau <- function(tau, df){
  df <- 
    df %>% 
      filter(!is.na(delta_ex)) %>% 
      mutate(
        T_param = ifelse(year < tau, FALSE, TRUE)
      )

  lm(delta_ex ~ T_param, data  = df)
}

# We can now run the above function, which creates a model, for a wide range of possible tau values
# This is the sequence listed and named tau_candidate, which is every whole number from 
# 1982 to 2015

# For each of these models, we can then get a measure of model fit. We're going to use AIC, though 
# could have used other measures. With AIC (and with BIC, and RMSE etc) better fit is indicated with 
# a lower value
grid_search_df <- tibble(
  tau_candidate = 1982:2015 # column of candidate values
) %>% 
  mutate(
    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df) # run the model with the value of tau in tau_candidate
  ) %>% 
  mutate(
    aic = map_dbl(model_alt, AIC)
  )

# Now we can see which proposed tau results in what fit value (AIC), initially just by graphing it.
grid_search_df %>% 
  ggplot(aes(tau_candidate, aic)) + 
  geom_point()

# So, in the above we found 2012 as tau minimised AIC. So, that's the best possible alternative model. 
# Let's run and label it as such
best_alt_model <- example_df %>% 
  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
  lm(delta_ex ~ T_param, data = .)

# Now we want to create the null model to compare it against (which just has an intercept term alone)
null_model <- example_df %>% 
  lm(delta_ex ~ 1, data = .)

# And finally, we want to compare the best_alt_model against the null model using an F test
anova(null_model, best_alt_model)
# We are interested in the P value here. If it's statistically significant then the best_alt_model does better
# (Is worth its additional complexity in terms of how much it makes the model fit) than the null model,
# So... for females, from birth, in England & Wales, there's evidence that there WAS a significant change in 
# average life expectancy improvement from 2012 onwards. (i.e. two years after austerity was introduced)

# We can manually repeat the same exercise for some additional subsets of the data (population, gender, starting age combinations). The below shows the same for males in England & Wales, from birth.

example_df <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979)

example_df <- 
  example_df %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex))

grid_search_df <- tibble(
  tau_candidate = 1982:2015
) %>% 
  mutate(
    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
  ) %>% 
  mutate(
    aic = map_dbl(model_alt, AIC)
  )

grid_search_df %>% 
  ggplot(aes(tau_candidate, aic)) + 
  geom_point()

best_alt_model <- example_df %>% 
  mutate( T_param = ifelse(year < 2013, FALSE, TRUE)) %>% 
  lm(delta_ex ~ T_param, data = .)

null_model <- example_df %>% 
  lm(delta_ex ~ 1, data = .)

anova(null_model, best_alt_model)

# This time 2013 is a slightly better candidate than 2012, hence 2013 being used in best_alt_model not 2012.

# Another example: France, males
###
example_df <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "FRATNP") %>% 
      filter(year >= 1979)

example_df <- 
  example_df %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex))

grid_search_df <- tibble(
  tau_candidate = 1982:2015
) %>% 
  mutate(
    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
  ) %>% 
  mutate(
    aic = map_dbl(model_alt, AIC)
  )

grid_search_df %>% 
  ggplot(aes(tau_candidate, aic)) + 
  geom_point()

best_alt_model <- example_df %>% 
  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
  lm(delta_ex ~ T_param, data = .)

summary(best_alt_model)

null_model <- example_df %>% 
  lm(delta_ex ~ 1, data = .)
summary(null_model)

anova(null_alt_model, best_alt_model)

# Here 2015 has the lowest value, but this might be an artefact of the data period covered. 
# We can also see that 2012 is the next best candidate.
```
